---
title: "Linking to and configuring RSeqAn"
author: "August Guang"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
    %\VignetteIndexEntry{Linking to and configuring RSeqAn}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

## Introduction

The reason [`RSeqAn`](https://bioconductor.org/packages/release/bioc/html/RSeqAn.html) was created was to allow for easy integration of the
SeqAn biological sequence analysis C++ library into R packages. This vignette describes how to link to [`RSeqAn`](https://bioconductor.org/packages/release/bioc/html/RSeqAn.html)
from another R package as well as how to configure RSeqAn for your own build system such as enabling `zlib` or `bzip2`.

## Linking

### Dependencies for linking

Prerequisites for linking to `RSeqAn` are:

 * Compiler needs to support C++14 standard. This is the default standard [from GCC6 on](https://en.wikipedia.org/wiki/C%2B%2B14#Compiler_support). You need to tell the build system to use C++14, either by modifying the `SystemRequirements` field of the `DESCRIPTION` file:
 ```
 SystemRequirements: C++14
 ```
 or (preferred) by specifying it in `src/Makevars`:
 ```
 CXX_STD = CXX14
 ```
 * [`Rcpp`](https://cran.r-project.org/package=Rcpp) needs to be installed and imported inside the `DESCRIPTION` file:
 ```
 Imports: Rcpp
 ```
 as well as specified in the `NAMESPACE` file:
 ```
 importFrom(Rcpp, sourceCpp)
 ```
 Note: If you generate your `NAMESPACE` with [`roxygen2`](https://cran.r-project.org/web/packages/roxygen2/index.html) then don't worry about the `NAMESPACE` file.
 
### Linking to RSeqAn

As long as the prerequisites are satisfied, then linking to `RSeqAn` is simple. Just put `RSeqAn` into the `Imports` field of the `DESCRIPTION` file as well, and then put
```
LinkingTo: Rcpp, RSeqAn
```
also in the `DESCRIPTION` file.

In C or C++ code, use `#include <seqan/$filename.h>` as usual, as well as `// [[Rcpp::depends(RSeqAn)]]` [as usual](http://dirk.eddelbuettel.com/code/rcpp/Rcpp-attributes.pdf). For an example, you can look at the [`qckitfastq`](https://bioconductor.org/packages/release/bioc/html/qckitfastq.html) package [source code](https://github.com/compbiocore/qckitfastq).

## Configuring RSeqAn

By default `SeqAn` and thus `RSeqAn` are not set up to make use of libraries like `zlib` and `bzip2` although it has the capabilities. In order to enable and set options for these libraries (assuming the libraries are installed), preprocessor flags for it should be set in `src/Makevars` (preferred) or using `Sys.setenv()`. As an example for enabling `zlib`:

 * In `src/Makevars`, write:
 ```
 PKG_CXXFLAGS=-DSEQAN_HAS_ZLIB
 ```
 * Using `Sys.setenv()`:
 ```
 Sys.setenv("PKG_CXXFLAGS"="-DSEQAN_HAS_ZLIB")
 ```

You can see other preprocessor defines that can be set at the [SeqAn documentation](https://seqan.readthedocs.io/en/master/Infrastructure/Use/CustomBuildSystem.html#preprocessor-defines-affecting-seqan).

### Example script

An example script using `Sys.setenv()` to set preprocessor defines that follows the SeqAn [SAM and BAM I/O tutorial](https://seqan.readthedocs.io/en/seqan-v2.0.2/Tutorial/SamAndBamIO.html) is below:

```{r setup}
Sys.setenv("PKG_CXXFLAGS"="-DSEQAN_HAS_ZLIB -std=c++14")
```

```{r bamio, engine='Rcpp'}
// [[Rcpp::depends(RSeqAn)]]

#include <seqan/bam_io.h>
#include <Rcpp.h>

using namespace Rcpp;

// [[Rcpp::export]]
int readBam()
{
    // test.bam is in vignettes folder
    seqan::CharString bamFileName = "toy.bam";

    // Open input file, BamFileIn can read SAM and BAM files.
    seqan::BamFileIn bamFileIn(toCString(bamFileName));

    // Open output file, BamFileOut accepts also an ostream and a format tag.
    // Note the usage of Rcout instead of std::cout
    seqan::BamFileOut bamFileOut(context(bamFileIn), Rcout, seqan::Sam());
    
    // Copy header.
    seqan::BamHeader header;
    seqan::readHeader(header, bamFileIn);
    seqan::writeHeader(bamFileOut, header);
    
    // Copy records.
    seqan::BamAlignmentRecord record;
    while (!atEnd(bamFileIn))
    {
        seqan::readRecord(record, bamFileIn);
        seqan::writeRecord(bamFileOut, record);
    }

    return 0;
}
```

```{r main}
readBam()
```